---
title: "Part 1"
author: "LeeLaven_240730653"
date: "2025-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Part 1A

## Aim to simulate random numbers from the target distribution with probability density function:

Algorithm Steps:

1\. Start at an initial value $X_0$.

2\. At each iteration $i$, propose a new candidate $x^* \sim \mathcal{N}(X_{i-1}, s^2)$ (random walk proposal).

3\. Compute the log acceptance ratio: $\log r = \log f(x^*) - \log f(X_{i-1})$

4\. Accept $x^*$ with probability $\min(1, r)$; otherwise stay at $X_{i-1}$.

5\. Repeat for $N$ iterations to obtain a chain of samples.

```{r target-density}
# Define the target density: f(x) = 0.5 * exp(-|x|)
DF <- function(x) {
  0.5 * exp(-abs(x))
}
```

## Random walk metropolis

X0: the initial starting value of the chain

N: the number of samples (iterations) to generate

s: the standard deviation of the proposal distribution

```{r}
set.seed(22)

rand_walk_metro <- function(X0, N, s) {
  samples <- numeric(N)  
  samples[1] <- X0       # Set starting value
  
  for (i in 2:N) {
    now <- samples[i - 1]
    x_star <- rnorm(1, mean = now, sd = s)              
    log_ratio <- log(DF(x_star)) - log(DF(now))         # Log acceptance ratio
    log_u <- log(runif(1))                              # Log of uniform random variable
    
    if (log_u < log_ratio) {
      samples[i] <- x_star   # Accept 
    } else {
      samples[i] <- now      # Reject: stay at current value
    }
  }
  
  return(samples)
}
```

# Parameters

```{r}
X0 <- 0
N <- 10001
s <- 1

result <- rand_walk_metro(X0, N, s)

cat("Number of samples:", length(result), "\n")
cat("Estimated Sample Mean:", mean(result), "\n")
cat("Estimated Sample Standard Deviation:", sd(result), "\n")

```

# Plot

The histogram of the generated samples is overlaid with:

-   The KDE — a smooth empirical approximation of the sample distribution
-   The True Target Density $f(x) = \frac{1}{2}e^{-|x|}$ — shown as a dashed curve for comparison

```{r}
result_df <- data.frame(x = result)

ggplot(result_df, aes(x = x)) +
  geom_histogram(aes(y = after_stat(density), fill = "Histogram"),
                 bins = 50, alpha = 0.6) +
  geom_density(aes(color = "Kernel Density Estimate"), linewidth = 1.2) +
  stat_function(fun = DF, aes(color = "True Density f(x)"),
                linewidth = 1.2, linetype = "dashed") +
  labs(title = "Metropolis-Hastings Samples vs Target Density",
       x = "x", y = "Density",
       fill = "Legend", color = "Legend") +
  scale_fill_manual(values = c("Histogram" = "lightblue")) +
  scale_color_manual(values = c("Kernel Density Estimate" = "purple",
                                "True Density f(x)" = "orange")) +


theme_gray()
```

## Part 1B

## Assess whether the algorithm has converged

where: $$M_j = \frac{1}{N}\sum_{i=1}^{N} x_i^{(j)}, \quad
  V_j = \frac{1}{N}\sum_{i=1}^{N}(x_i^{(j)} - M_j)^2$$ $$W = \frac{1}{J}\sum_{j=1}^{J} V_j, \quad
  M = \frac{1}{J}\sum_{j=1}^{J} M_j, \quad
  B = \frac{1}{J}\sum_{j=1}^{J}(M_j - M)^2$$

```{r}
R_hat <- function(chains) {
  N  <- length(chains[[1]])
  Mj <- sapply(chains, mean)
  Vj <- sapply(chains, function(chain) mean((chain - mean(chain))^2))

  W <- mean(Vj)           # Within-chain variance
  M <- mean(Mj)           # Grand mean
  B <- mean((Mj - M)^2)  # Between-chain variance

  return(sqrt((B + W) / W))
}

```

#to run J chains from dispersed starting points

```{r}
run_chains <- function(N, s, J, initial) {
  chains <- list()
  for (j in 1:J) {
    chains[[j]] <- rand_walk_metro(initial[j], N, s)
  }
  return(chains)
}
```

## R-hat for s= 0.001

```{r}
set.seed(22)
s       <- 0.001
N       <- 2000
J       <- 4
initial <- c(-10, -5, 5, 10)

chains <- run_chains(N, s, J, initial)
cat("R-hat for s = 0.001:", R_hat(chains), "\n")
```

## R-hat Over a Grid of s Values

```{r}
grid          <- seq(0.001, 1, length.out = 1000)
R_hat_results <- numeric(length(grid))

for (i in seq_along(grid)) {
  chains          <- run_chains(N, grid[i], J, initial)
  R_hat_results[i] <- R_hat(chains)
}

```

## Plot

```{r}
results_df <- data.frame(s = grid, R_hat = R_hat_results)

ggplot(results_df, aes(x = s, y = R_hat)) +
  geom_line(aes(color = "R-hat"), linewidth = 1) +
  geom_hline(aes(yintercept = 1, color = "Threshold (R-hat = 1)"),
             linetype = "dashed", linewidth = 0.8) +
  labs(title = expression("R-hat vs Proposal Standard Deviation (s)"),
       x = "s (Proposal Standard Deviation)",
       y = expression(hat(R)),
       color = "Legend") +
  scale_color_manual(values = c("R-hat" = "purple",
                                "Threshold (R-hat = 1)" = "hotpink")) +
theme_gray()

```
